---
title: "sample"
author: "Amal Tom"
date: "2024-11-10"
output:
  pdf_document: default
  html_document: default
---

```{r}
options(repos = c(CRAN = "https://cran.rstudio.com"))

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## HR Data Analytics
The below document consist of exploration and analysis of a HR dataset.
Contains various data exploration methods,visualizations to understand the data in a better way.

## Navigate to the bottom for inferences.



Loading the dataset

```{r}
data=read.csv("E:/RCSS/SEM 3/R/hrdata.csv")
data
data1=data
```
# Data Explorations:-

```{r}
colnames(data)
```
```{r}
age=unique(data$Age)
degree=unique(data$Degree)
degree_100=unique(data$Degree.Percentage)
experience=unique(data$experience)
department=unique(data$Department)
salary=unique(data$Salary)

age
degree
degree_100
experience
department
salary
```

# Finding total salary contribution from each department

```{r}
dep_salary <- tapply(data$Salary, data$Department, sum)
dep_salary

```
# Merging 'marketing' with 'Marketing'
```{r}
data1$Department[data1$Department == "marketing"] <- "Marketing"
dep_salary <- tapply(data1$Salary, data1$Department, sum)
dep_salary

```
# Inference 1:Most of the salary is being utilised by the Marketing department


#Representing through Bar Plot
```{r}
values <- c(A = 10, B = 15, C = 7)
barplot(dep_salary, main = "Bar Plot Example", xlab = "Categories", ylab = "Counts", col = "skyblue")

```


# Number of employees across each departments:-
```{r}
# Count occurrences of each unique value in 'Department' column
dep_count<-table(data1$Department)
dep_count
```
# Inference 2:Most of the employees are under the Marketing department


# Average salary of employees across different branches:-
```{r}
avg=dep_salary/dep_count
avg
```
# Inference 3:Average salary of employees working under 'Project' department is the highest


# Salary Range across different branches:-
```{r}
salary_range_by_dept <- tapply(data1$Salary, data1$Department, range)
print(salary_range_by_dept)

```
# Inference 4: The Marketing department has the the lowest and highest allotted salary(Range:11000-19000)



# Converting categorical 'degree' to numerical values-'Degree':-
```{r}
data1$degree <- as.numeric(as.factor(data1$Degree))
Degree<-unique(data1$Degree)
data1
```
`

# Creating correlation matrix, to find dependency between attributes:-
```{r}
cor_matrix <- cor(data1[, c("Salary","experience","Degree.Percentage","Year.of.Joining","Age","degree")])
print(cor_matrix)

```


# Heatmap representation of the correlation matrix:-
```{r}
install.packages("corrplot")
library(corrplot)
corrplot(cor_matrix, 
         method = "color", 
         addCoef.col = "black", 
         number.cex = 0.7, 
         tl.cex = 0.8, 
         main = "Correlation Matrix Heatmap (corrplot)")

```
# Inference 5: Experience has a greater influence on salary compared with other attributes.



# Finding the age range and number of employees across different departments:- 
```{r}
library(dplyr)
age_range_by_dept <- data1 %>%
  group_by(Department) %>%
  summarise(
    min_age = min(Age, na.rm = TRUE),
    max_age = max(Age, na.rm = TRUE),
    mean_age = mean(Age, na.rm = TRUE),
    median_age = median(Age, na.rm = TRUE),
    employee_count = n()
  )

print(age_range_by_dept)

```
# Inference 6:The employees under Project department are very much senior to the other department employees




```{r}
project_legal_details <- data1 %>%
  filter(Department %in% c("Project", "legal","IT"))
print(project_legal_details)

```
# Inference 7:- people working under the departments-Project,Legal and IT contributes lesser number of empoyees.



```{r}

salary_intervals <- cut(data1$Salary, 
                        breaks = seq(11000, 91000, by = 10000), 
                        include.lowest = TRUE, 
                        right = FALSE,   # If you want intervals like [18000, 29000)
                        labels = c("11k-21k", "21k-31k", "31k-41k", "41k-51k", "51k-61k", "61k-71k", "71k-81k","81k-91k"))
data1$SalaryInterval <- salary_intervals
data1
```



# Representing Experience with respect to Salary
```{r}
plot(data1$experience, data1$Salary,
     main = "Marketing Experience vs Salary",
     xlab = "Experience (Years)",
     ylab = "Salary",
     pch = 19, col = "blue")

```
# Inference 8:-Even highest experienced employees have salary range from 11000 to 91000



# Exploring employees having experience for more than 25 years and below median salary
```{r}
library(ggplot2)
library(dplyr)
employees_filtered <- data1 %>%
  filter(experience > 25 & Salary < 45000)
ggplot(employees_filtered, aes(x = Degree)) +
  geom_bar(fill = "skyblue") +
  labs(title = "Employees with >25 years Experience and Salary < 60000 by Degree",
       x = "Degree",
       y = "Number of Employees") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  theme_minimal()  

```
# Inference 9:- Experience doesn't count for Salary hike with lower Educational qualifications like sslc, PDC



# Grouping employees by SalaryInterval and Experience and their respective counts
```{r}
experience_by_salary_interval <- data1 %>%
  group_by(SalaryInterval, experience,Degree,Degree.Percentage
) %>%
  summarise(Count = n(), .groups = "drop")   # .groups = "drop" removes the grouping structure

print(experience_by_salary_interval)


```


# Number of employees under different Salary scale
```{r}
s<-table(experience_by_salary_interval$SalaryInterval)
s
```
# Inference 10:-Most number of employees are having starting salaries.
# Inference 11:Number of employees decreases as salary increases.



# Finding the distribution of degree percentage across Salary Scale
```{r}
library(dplyr)

experience_degree_summary <- data1 %>%
  group_by(SalaryInterval, experience, Degree.Percentage) %>%
  summarise(Count = n(), .groups = "drop")
print(experience_degree_summary)

ggplot(experience_degree_summary, aes(x = SalaryInterval, y = Count, fill = Degree.Percentage)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Degree Percentage Distribution Across Salary Intervals",
       x = "Salary Interval",
       y = "Count of Degree Percentages") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  
  theme_minimal()


```
# Inference 12:-Employees with lowest degree percentage are having basic Salary(11k-21k)
# Inference 13: Employees with highest degree percentage has highest salary(81k-91k)


# Creating a stacked Bar Chart to observe degree distributions across Salaries
```{r}

library(ggplot2)
degree_salary_summary <- data1 %>%
  group_by(SalaryInterval, Degree) %>%
  summarise(Count = n(), .groups = "drop")
print(degree_salary_summary)

ggplot(degree_salary_summary, aes(x = SalaryInterval, y = Count, fill = Degree)) +
  geom_bar(stat = "identity", position = "stack") +  # Stack bars by Degree
  labs(title = "Distribution of Degree Types Across Salary Intervals",
       x = "Salary Interval",
       y = "Count of Degrees") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels for readability
  theme_minimal()  # Clean theme
```
# Inference 14:-SSLC and PDC degree holders have lowest pay scale
# Inference 15:-Degree doesn't have a significant relationship with salary


```{r}
experience_by_salary_interval <- data1 %>%
  group_by(SalaryInterval, experience, Degree, Degree.Percentage) %>%
  summarise(Count = n(), .groups = "drop")
print(experience_by_salary_interval)

ggplot(experience_by_salary_interval, aes(x = SalaryInterval, y = Count, fill = experience)) +
  geom_bar(stat = "identity", position = "stack") +  # Create a stacked bar plot
  labs(title = "Experience Levels Across Salary Intervals",
       x = "Salary Interval",
       y = "Count of Experience Levels") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  
  theme_minimal()

```
# Inference 16:-Employees with 30+ year of experience has the highest Salary(81k-91k)
# Inference 17:-Employees with few years of experience has basic salaries.


# INFERENCES:-
1: Most of the salary is being utilised by the Marketing department(1228000).

2: More than half of the total employees are under the Marketing department-58%.(35 out of 60).

3: Average salary of employees working under 'Project' department is the highest(49666.67).

4: The Marketing department has the lowest and highest allotted salary (Range:11000-19000).

5: Experience has a greater influence on salary compared with other attributes.

6: The employees under Project department are very much senior to the other department employees.

7: People working under the departments-Project, Legal and IT contributes lesser number of         employees(3,1,3).

8: Even highest experienced employees have salary range from 11000 to 91000.

9: Experience doesn't count for Salary hike with lower educational qualifications like SSLC, PDC.

10: Most number of employees are having starting salaries.

11: Number of employees decreases as salary increases.

12: Employees with lowest degree percentage are having basic Salary(11k-21k).

13: Employees with highest degree percentage has highest salary(81k-91k).

14: SSLC and PDC degree holders have lowest pay scale.

15: Degree doesn't have a significant relationship with salary.

16: Employees with 30+ year of experience has the highest salary(81k-91k).

17: Employees with few years of experience has basic salaries.

# Summary
By analysing the given dataset it is clear that employees with basic degree qualification tends to have increase in Salary as their experience increases.





